Mixture Models for Single-Cell Assays with Application to Vaccine Studies

### 2013-05-29 MIMOSA.tex
Fully revised and shortened. Ready to submit

\begin{abstract}
{Blood and tissue are composed of many functionally distinct cell subsets. In immunological studies, these can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify biomarkers (or combinations of biomarkers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after stimulation), where expression is defined as the proportion of cells expressing that biomarker (or biomarker combination) in the cell subset(s) of interest.
Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential biomarker expression using single-cell assays. Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against frequentist approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods. Additional simulations show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.}{Expectations-Maximization, Markov Chain Monte Carlo,
Marginal Likelihood, Bayesian Modeling, Hierarchical Modeling,
Immunology, Flow Cytometry, Single-Cell Gene Expression, MIMOSA}
\end{abstract}

\clearpage
\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them.
This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures.
For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins.
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM).
Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular biomarkers (up to 17 biomarkers).
More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e. DNA content, cell--surface or intracellular protein levels, protein phosphorylation, RNA--level gene expression, etc.) per cell~\citep{Bendall:2011wf}.
Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}.
While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}.
This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data.
Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general.
For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
These assays typically measure multiple biomarkers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

@@

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. The antigen--specific T-cell subpopulations (i.e. the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells (on the order of 0.1\% for HIV vaccine studies like the one shown here). The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
Individual cells are labelled using fluorescently conjugated antibodies against cell--surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations can be detected.
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.

@@

\section{Differential expression with one biomarker}
\label{s:DEone}
Datasets like the ones presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations. By univariate, we mean that we have only one positive cell subset. This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ and/or B$+$ (\textit{vs.} A$-$/B$-$). Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$). In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.

@@

\subsection{Priors}
\label{ss:priors}
Our model shares information across all subjects using exchangeable Beta priors on the unknown proportions, as follows:
 \begin{align*}
(p^{(u)}_{i} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha^{(u)}, \beta^{(u)})\\
(p^{(u)}_{i}|z_{i}=1) \sim \mathrm{Beta}(\alpha^{(u)},
\beta^{(u)}) \quad\mathrm{and}\quad (p^{(s)}_{i} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha^{(s)},\beta^{(s)}),
 \end{align*}
where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, \textit{i.e.}, ${\cal M}_1$ is true, and zero otherwise, and $\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}$ are unknown hyper-parameters shared across all subjects.
Note that the parameters $\alpha^{(u)} \text{ and } \beta^{(u)}$ are explicitly shared across the two models, whereas $\alpha^{(s)} \text{ and } \beta^{(s)}$ are only present in the alternative model.
Finally, we assume that $z_i\sim \mathrm{Be}(w)$ are independent draws from a Bernoulli distribution with probability $w$, where $w$ represents the (unknown) proportion of responders.
It follows that marginally, \textit{i.e.}, after integrating $z_i$, the $p^{(u)}_{i}$ and $p^{(s)}_{i}$ are then jointly distributed as a mixture of a one dimensional Beta distribution and a product of two Beta distributions (with a possible constraint), with mixing parameter $w$.
Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}, w)$ can be estimated in an Empirical-Bayes fashion using Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}.
As an alternative, we also describe a fully Bayesian model, where the hyperparameters $\alpha^{(u)}, \beta^{(u)}$, $\alpha^{(s)},\text{ and }\beta^{(s)}$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
In this case, all parameters will be estimated via a Markov chain Monte Carlo algorithm as described in Section \ref{s:estimation}.

@@

and
\begin{equation}
\mathrm{L}_1(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\mathbf{y}_i)
=\binom{N^{(u)}_{i}}{n^{(u)}_{i}} \binom{N^{(s)}_{i}}{n^{(s)}_{i}}\cdot\frac{\mathrm{B}(n^{(u)}_{i}+\alpha^{(u)},N^{(u)}_{i}-n^{(u)}_{i}+\beta^{(u)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}\cdot \frac{\mathrm{B}(n^{(s)}_{i}+\alpha^{(s)},N^{(s)}_{i}-n^{(s)}_{i}+\beta^{(s)})}{\mathrm{B}(\alpha^{(s)},\beta^{(s)})}
\label{model2:unconstrained}
\end{equation}
Above, $\mathrm{B}$ is the Beta function. Assuming that the missing data, $z_i, i=1,\dots,I$, are known, we define the complete data log-likelihood:
\begin{equation}
\begin{split}
l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha^{(u)}, \beta^{(u)}|\mathbf{y}_i) +(1-z_i) l_1(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)}|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w),\label{eq:cll}
\end{split}
\end{equation}
where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}, w)$ is the vector of parameters to be estimated. In the one-sided case, the alternative prior specification must satisfy the constraint $p^{(s)}>p^{(u)}$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed-form but can easily be estimated. All calculations for the one-sided case are described in Appendix A of the Supplementary Material.

@@

\noindent\textbf{MCMC algorithm}\\
We generated realizations from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs samplings.
Details about the algorithms are given in Appendix B of the Supplementary Material. We used the
method of \cite{Raftery:1992vp} and \cite{Raftery:1996ws} to determine
the number of iterations, based on a short pilot run of the
sampler. For each dataset presented here, we calculated that no more
than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. To leave some margin, we used 200,000 iterations after 50,000 burn-in iterations for each dataset explored here.

\section{Results}
\label{s:results}
In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data. We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Newton2004} where a false discovery (for the ICS data) is a day 0 sample or placebo recipient (expected non--responder) that is incorrectly identified as a responder.

\subsection{ICS}
In the ICS data we considered observations from vaccinees and placebos at the day 0 time point and day 182 time point as true negatives, and observations from vaccinees at the day 182 time point as true positives. This assumption was required, since we do not have independent, external measurement of the true response for the day 182 vaccinees. Under these assumptions, the MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected.
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F). Here we present the results based on IL2 and IFN-$\gamma$ alone as well as the subset expressing IL2 or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}. These results are consistent for other cytokines and cytokine combinations, including higher order combinations  (see Supplementary Figure 1).  We also evaluated,through simulation, what would happen to the ROC curves when our assumptions about true responders and non-responders did not hold (\textit{i.e.,} when not all day 182 vaccinees are true responders, Supplementary Figure 2), and found that we are potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182.

@@

We verified the assumption of a common distribution for the $p^{(u)}_{i}$
across subjects by examining the empirical estimates of $p^{(u)}_{i}$ and the
density of the Beta distribution with parameters $\hat{\alpha}^{(u)},
\hat{\beta}^{(u)}$ fitted to ENV-1-PTEG stimulated
CD4+/IFN--$\gamma$+ T--cell ICS data (Supplementary Figure 3). The effect of information sharing on the sensitivity and specificity of the model was also investigated by varying the number of subjects in simulations (Supplementary Figure 4). Our model performed well when as few as 20 subjects were included, although the observed false discovery rate was below the nominal rate with so few subjects.

\subsection{Single-cell gene expression}
We applied the MIMOSA model to a Fluidigm single-cell gene expression data set. We used the two-sided MIMOSA model because genes could be regulated upward or downward upon stimulation. In order to detect stimulation specific changes of expression, we fit our model to each gene within each stimulation. The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A). These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A) and closely match the empirical estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B). As a comparison, we also fit the model across all stimulations (Figure~\ref{fig:fluidigm} C). Although this leads to a sensible clustering by stimulation, the observed patterns do not agree as well with the empirical differences. A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust p-values (Figure~\ref{fig:fluidigm} D) does not reveal stimulation-specific patterns, with the exception of CMV pp65 nlv5 stimulation). At an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes. The 13 genes identified by Fisher's method were a subset of those identified by MIMOSA.

@@

\subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
As a proof-of-concept, we applied our multivariate MIMOSA model for two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example, $K=2$, and we have four possible combinations.
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the sum of the counts in the 3 different positive subsets (Figure~\ref{fig:polyfunctionality} A,B). The number of CCR7-GZML+ cells increases upon stimulation, while the number of CCR7+GZMK- cells decreases. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all possible polyfunctional cell populations (\textit{i.e.}, cells expressing IL2$+$ or IFN-$\gamma+$), would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. The multivariate MIMOSA model tests all polyfunctional cell subpopulations simultaneously, and detects significant differences between stimulated and unstimulated conditions in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, black labels). In contrast, marginalizing over these cell populations, no difference is apparent in any of the samples (Figure~\ref{fig:polyfunctionality} B). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.

@@

Since the Fluidigm data set has a limited number of observations (100 cells and 16 samples), we could not look at more than two biomarkers at once. Therefore, we performed simulations in eight dimensions to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test and the likelihood ratio test on the resulting 2x8 tables (Figure~\ref{fig:mvsimulations} A-C). These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model better fits the data than the competing standard approaches tested (Figure~\ref{fig:mvsimulations} B).

@@

Appendix A, B, and C and Supplementary Figures 1--6 referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:results} and \ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}

>>>>

\begin{abstract}
{Blood and tissue are composed of many functionally distinct cell subsets. In immunological studies, these can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify biomarkers (or combinations of biomarkers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after stimulation), where expression is defined as the proportion of cells expressing that biomarker (or biomarker combination) in the cell subset(s) of interest.
Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential biomarker expression using single-cell assays. Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods. Additional simulations show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.}{Expectations-Maximization, Markov Chain Monte Carlo,
Marginal Likelihood, Bayesian Modeling, Hierarchical Modeling,
Immunology, Flow Cytometry, Single-Cell Gene Expression, MIMOSA}
\end{abstract}

\clearpage
\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them.
This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures.
For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins.
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM).
Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular biomarkers (up to 17 biomarkers).
More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e. DNA content, cell-surface or intracellular protein levels, protein phosphorylation, RNA--level gene expression, etc.) per cell~\citep{Bendall:2011wf}.
Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}.
While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}.
This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data.
Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general.
For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
These assays typically measure multiple biomarkers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

@@

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. The antigen--specific T-cell subpopulations (i.e. the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells (on the order of 0.1\% for HIV vaccine studies like the one shown here). The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
Individual cells are labelled using fluorescently conjugated antibodies against cell-surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations can be detected.
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.

@@

\section{Differential expression with one biomarker}
\label{s:DEone}
Datasets like the ones presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations. By univariate, we mean that we have only one positive cell subset. This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ or B$+$ (\textit{vs.} A$-$/B$-$). Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$). In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.

@@

\subsection{Priors}
\label{ss:priors}
Our model shares information across all subjects using exchangeable Beta priors on the unknown proportions, as follows:
 \begin{align*}
(p^{(u)}_{i} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha^{(u)}, \beta^{(u)})\\
(p^{(u)}_{i}|z_{i}=1) \sim \mathrm{Beta}(\alpha^{(u)},
\beta^{(u)}) \quad\mathrm{and}\quad (p^{(s)}_{i} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha^{(s)},\beta^{(s)}),
 \end{align*}
where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, \textit{i.e.}, ${\cal M}_1$ is true, and zero otherwise, and $\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}$ are unknown hyper-parameters shared across all subjects.
Note that the parameters $\alpha^{(u)} \text{ and } \beta^{(u)}$ are explicitly shared across the two models, whereas $\alpha^{(s)} \text{ and } \beta^{(s)}$ are only present in the alternative model.
Finally, we assume that $z_i\sim \mathrm{Be}(w)$ are independent draws from a Bernoulli distribution with probability $w$, where $w$ represents the (unknown) proportion of responders.
It follows that marginally, \textit{i.e.}, after integrating $z_i$, the $p^{(u)}_{i}$ and $p^{(s)}_{i}$ are then jointly distributed as a mixture of a one dimensional Beta distribution and a product of two Beta distributions (with a possible constraint), with mixing parameter $w$.
Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}, w)$ can be estimated in an Empirical-Bayes fashion using an Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}.
As an alternative, we also describe a fully Bayesian model, where the hyperparameters $\alpha^{(u)}, \beta^{(u)}$, $\alpha^{(s)},\text{ and }\beta^{(s)}$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
In this case, all parameters will be estimated via a Markov chain Monte Carlo algorithm as described in Section \ref{s:estimation}.

@@

and
\begin{equation}
\mathrm{L}_1(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\mathbf{y}_i)
=\binom{N^{(u)}_{i}}{n^{(u)}_{i}} \binom{N^{(s)}_{i}}{n^{(s)}_{i}}\cdot\frac{\mathrm{B}(n^{(u)}_{i}+\alpha^{(u)},N^{(u)}_{i}-n^{(u)}_{i}+\beta^{(u)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}\cdot \frac{\mathrm{B}(n^{(s)}_{i}+\alpha^{(s)},N^{(s)}_{i}-n^{(s)}_{i}+\beta^{(s)})}{\mathrm{B}(\alpha^{(s)},\beta^{(s)})}
\label{model2:unconstrained}
\end{equation}
Above, $\mathrm{B}$ is the Beta function. Assuming that the missing data, $z_i,i=1,\dots,I$, are known, we define the complete data log-likelihood:
\begin{equation}
\begin{split}
l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha^{(u)}, \beta^{(u)}|\mathbf{y}_i) +(1-z_i) l_1(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)}|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w),\label{eq:cll}
\end{split}
\end{equation}
where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}, w)$ is the vector of parameters to be estimated. In the one-sided case, the alternative prior specification must satisfy the constraint $p^{(s)}>p^{(u)}$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed-form but can easily be estimated. All calculations for the one-sided case are described in Appendix A of the Supplementary Material.

@@

\noindent\textbf{MCMC algorithm}\\
We generated realizations from the posterior distribution via MCMC algorithms~\citep{Walter1996}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs samplings.
Details about the algorithms are given in Appendix B of the Supplementary Material. We used the
method of \cite{Raftery:1992vp} and Raftery \citep{Walter1996} to determine
the number of iterations, based on a short pilot run of the
sampler. For each dataset presented here, we calculated that no more
than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. To leave some margin, we used 200,000 iterations after 50,000 burn-in iterations for each dataset explored here.

\section{Results}
\label{s:results}
In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data. We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set where a false discovery (for the ICS data) is a day 0 sample or placebo recipient (expected non--responder) that is incorrectly identified as a responder. For MIMOSA, the FDR was computed using the approach of~\citet{Newton2004} whereas for all other methods FDR control was done using the approach of~\citet{Benjamini1995}.

\subsection{ICS}
In the ICS data we considered observations from vaccinees and placebos at the day 0 time point and day 182 time point as true negatives, and observations from vaccinees at the day 182 time point as true positives. This assumption was required, since we do not have independent, external measurement of the true response for the day 182 vaccinees. Under these assumptions, the MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected.
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F). Here we present the results based on IL2 and IFN-$\gamma$ alone as well as the subset expressing IL2 or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}. These results are consistent for other cytokines and cytokine combinations, including higher order combinations  (see Supplementary Figure 1).  We also evaluated, through simulation, what would happen to the ROC curves when our assumptions about true responders and non-responders did not hold (\textit{i.e.,} when not all day 182 vaccinees are true responders, Supplementary Figure 2), and found that we are potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182. Nonetheless, the ranking of the methods is consistent and the conclusions from the ROC analysis still hold.

@@

We verified the assumption of a common distribution for the $p^{(u)}_{i}$
across subjects by examining the empirical estimates of $p^{(u)}_{i}$ and the
density of the Beta distribution with parameters $\hat{\alpha}^{(u)},
\hat{\beta}^{(u)}$ fitted to ENV-1-PTEG stimulated
CD4+/IFN--$\gamma$+ T--cell ICS data (Supplementary Figure 3). The effect of information sharing on the sensitivity and specificity of the model was also investigated by varying the number of subjects in simulations (Supplementary Figure 4). Our model performed well when as few as 20 subjects were included, although the nominal false discovery rate overestimated observed false discovery rate.

\subsection{Single-cell gene expression}
We applied the MIMOSA model to a Fluidigm single-cell gene expression data set. We used the two-sided MIMOSA model because genes could be regulated upward or downward upon stimulation. In order to detect stimulation specific changes of expression, we fit our model to each gene within each stimulation. The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A). These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A) and closely match the empirical estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B). As a comparison, we also fit the model across all stimulations (Figure~\ref{fig:fluidigm} C). Although this leads to a sensible clustering by stimulation, the observed patterns do not agree as well with the empirical differences. A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust q-values (Figure~\ref{fig:fluidigm} D) does not reveal stimulation-specific patterns, with the exception of CMV pp65 nlv5 stimulation. At an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes. The 13 genes identified by Fisher's method were a subset of those identified by MIMOSA. These results suggest that fitting a stimulation--specific model is the right approach since information is shared across subjects that are expected to exhibit similar behavior.

@@

\subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
As a proof-of-concept, we applied our multivariate MIMOSA model for two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example, $K=2$, and we have four possible combinations.
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the $\log_2$ counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the sum of the counts in the 3 different positive subsets (Figure~\ref{fig:polyfunctionality} A,B). The number of CCR7-GZML+ cells increases upon stimulation, while the number of CCR7+GZMK- cells decreases. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all possible polyfunctional cell populations (\textit{i.e.}, cells expressing CCF7$+$ or GZMK$+$), would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. The multivariate MIMOSA model tests all polyfunctional cell subpopulations simultaneously, and detects significant differences between stimulated and unstimulated conditions in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, dark labels). In contrast, marginalizing over these cell populations, no difference is apparent in any of the samples (Figure~\ref{fig:polyfunctionality} B). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.

@@

Since the Fluidigm data set has a limited number of observations (100 cells and 16 samples), we could not look at more than two biomarkers at once. Therefore, we performed simulations in eight dimensions to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test and the likelihood ratio test on the resulting 2x8 tables (Supplementary Figure 6 A,B). These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model better fits the data than the competing standard approaches tested (Supplementary Figure 6 B).

@@

Appendix A, B, and C and Supplementary Figures 1--7 referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:results} and \ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}
